A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-11 tex/sec5-synthesis.tex
[ tex ] Revisions here and there

\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
This section focuses on defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
We start with \Cref{sec:mode-correctness} by defining mode-correctness and showing the uniqueness of synthesized types.
This uniqueness means that any two synthesised types for the same raw term $t$ under the same context $\Gamma$ have to be equal.
It will be used especially in \Cref{subsec:bidirectional-synthesis-checking} for the proof of the decidability of bidirectional type synthesis and checking.
Then, we conclude this section with the trichotomy on raw terms in \Cref{subsec:trichotomy}.

\subsection{Mode Correctness}\label{sec:mode-correctness}
As \citet{Dunfield2021} outlined, mode-correctness for a bidirectional typing rule means that 
\begin{enumerate*}
\item each `input' type variable in a premise must be an `output` variable in `earlier' premises or provided by the conclusion, if the rule is checking;
\item each `output' type variable in the conclusion should be some `output' variable in premises, if the rule is synthesis.
\end{enumerate*}
Here `input' variables refers to variables in an extension context and in the type of a premise in the checking mode.
It is important to note that the order of premises in a bidirectional typing rule also matters, since synthesis type variables are instantiated incrementally during type synthesis.

@@

Now let us define mode-correctness rigorously.
As we have outlined, the condition of mode-correctness for a synthesis rule is different from that of a checking rule, and the argument order also matters.
Defining the condition directly for a rule, and thus in our setting for an operation, can be somewhat intricate.
Instead, we choose to define the conditions for the argument list---more specifically, triples $\biargvec$ of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, for an operation, and subsequently for a signature.
We also need some auxiliary definitions for the subset of variables of a type and of a context and the set of variables that have been synthesised:
\begin{definition}
  The finite subset\footnote{%
  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
  However, for our purposes, the choice among these definitions is not a matter of concern.}
  of \emph{(free) variables} of a type $A$ is denoted by $\fv(A)$.
  The subset $\fv(\Gamma)$ of variables in a context $\Gamma$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Gamma, A) = \fv(\Gamma) \cup \fv(A)$.
  For an argument list $\biargs$, define its set of \emph{synthesis type variables} inductively as 
  \begin{align}
    \label{eq:synvar1}\synvar(\cdot)                   & = \emptyset  \\
    \label{eq:synvar2}\synvar(\biargs, \chkbiarg[n+1]) & = \phantom{\fv(A_{n+1}) \cup {}} \synvar(\biargs) \\
    \label{eq:synvar3}\synvar(\biargs, \synbiarg[n+1]) & = \fv(A_{n+1}) \cup           \synvar(\biargs).
  \end{align}
\end{definition}
This subset contains type variables of a synthesis argument and they are exactly those type variables that would be synthesised during type synthesis.

\begin{definition}\label{def:mode-correctness-args}
  The \emph{mode-correctness} $\MCas\left(\biargvec\right)$ for an argument list $\biargs$ with respect to a subset $S$ of $\Xi$ is defined as
  \begin{align}
    \label{eq:MC1} \MCas( \cdot ) & = \top \\
    \label{eq:MC3} \MCas\left(\biargvec, \chkbiarg[n]\right)
                                  & = \fv(\Delta, A_n) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land \MCas\left(\biargvec\right) \\
    \label{eq:MC2} \MCas\left(\biargvec, \synbiarg[n]\right) 
                                  & = \phantom{, A_n} \fv(\Delta) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land  \MCas\left(\biargvec\right)
  \end{align}
  where \eqref{eq:MC1} means an empty list is always mode-correct.
\end{definition}
This definition encapsulates the idea that every `input' type variable, possibly deriving from a context $\Delta$ or a checking argument $A_n$, must be an `output' variable from $\synvar(\biargvec)$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
This condition must be met for every tail of the argument list as well to ensure that `output' variables accessible at each argument position are from preceding arguments only, hence an inductive definition.
\begin{definition}\label{def:mode-correctness}
  An operation $\biop$ is \emph{mode-correct} if 
  \begin{enumerate}
    \item either $d$ is $\chk$, its argument list is mode-correct w.r.t.\ $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\biargvec)$ contains every inhabitant of $\Xi$;
    \item or $d$ is $\syn$, its argument list is mode-correct w.r.t.\ $\emptyset$, and the finite subset $\synvar(\biargvec)$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
  \end{enumerate}
  A bidirectional binding signature $\Omega$ is \emph{mode-correct} if its operations are all mode-correct.
\end{definition}
For a checking operation, an `input' variable of an argument could be derived from $A_0$, as these are known during type checking as an input.
Since every inhabitant of $\Xi$ can be located in either $A_0$ or synthesis variables, we can determine a concrete type for each inhabitant of $\Xi$ during type synthesis.
On the other hand, for a synthesis operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to $\emptyset$.
Also, the set of synthesis variables alone should include every type variable in $\Xi$ and particularly in $A_n$.
It is also easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) in \Cref{ex:signature-simply-typed-lambda} for simply typed $\lambda$-calculus is mode-correct according to our definition.

\begin{remark}
  Mode-correctness is fundamentally a condition for typing \emph{rules} themselves, not for derivations.
  Thus, this property cannot be formalised without treating rules as some mathematical object, such as the notion of bidirectional binding signature presented in \Cref{sec:defs}.
  This contrasts with the properties in \Cref{sec:pre-synthesis}, which can still be specified for individual systems even in the absence of a generic definition of bidirectional type systems.
\end{remark}

Now, we set out to show the uniqueness of the synthesised types for a mode-correct bidirectional type system.
For a specific system, its proof is typically a straightforward induction.
However, given the mode-correctness---which incorporates an inductive definition on the argument list---we establish our proof through induction, both on typing derivations and on the argument list:
\begin{lemma}[Uniqueness of the Synthesised Types]\label{thm:unique-syn}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
    \quad\text{and}\quad
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
  \]
  for the same term $t$ must be equal, i.e.\ $A = B$.
\end{lemma}
\begin{proof}%[Proof of \Cref{thm:unique-syn}]
  We prove the statement by induction on derivations $d_1$ and $d_2$ for $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ and $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$.
  Our system is syntax-directed, so $d_1$ and $d_2$ must be derived from the same rule: 
  \begin{itemize}
    \item $\SynRule{Var}$ follows from that each variable as a raw term refers to the same variable in its context.
    \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotate A$ in question.
    \item $\Rule{Op}$: Recall that a derivation of\/ $\Gamma |- \tmOpts \syn A$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types.
      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$ and $\rho_2$ of $d_1$ and $d_2$, respectively, agree on variables in $\synvar(\biargs)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
      We prove it by induction on the argument list:
      \begin{enumerate}
        \item For the empty list, the statement is vacuously true by \Cref{eq:synvar1}.
        \item If $\dir{d_{i+1}}$ is checking, then the statement holds for \Cref{eq:synvar2} by induction hypothesis.
        \item If $\dir{d_{i+1}}$ is synthesis, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by mode-correctness \Cref{eq:MC2} and induction hypothesis (of the list).
          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$, as required for the case \Cref{eq:synvar3} in the definition of $\synvar$.
      \end{enumerate}
  \end{itemize}
\end{proof}

@@

We are here for the main technical contribution of this paper.

@@

The interesting part of the theorem is the case for the $\Rule{Op}$ rule and it is rather complex.
Here we shall give its insight first instead of jumping into the details.
Recall that a typing derivation for $\tmOpts$ contains a substitution $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$.
The goal of type synthesis is exactly to define such a substitution $\rho$, so we have to start with an `accumulating' substitution: a substitution $\rho_0$ that is partially defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere.
By mode-correctness, the accumulating substitution~$\rho_i$ will be defined on enough synthesis variables so that type synthesis or checking can be performed on $t_{i}$ with the context $\Gamma, \vars{x}_{i} : \simsub{\Delta_{i}}{\rho_{i}}$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
If we visit a synthesis argument $\synbiarg[i+1]$, then we may extend the domain of $\rho_i$ to include the synthesis variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and thereby \emph{extend} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$ with the unifier.
Then, if we go through every $t_i$ successfully, we will have a total substitution $\rho_n$ by mode-correctness and a derivation of $\Gamma, \vars{x}_i : \Delta_i |-_{\Sigma, \Omega} t_i :^{\dir{d_i}} \simsub{A}{\rho_n}$ for each subterm $t_i$.

@@

\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis-checking}}]
  We prove this statement by induction on the mode derivation\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$.
  The two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness.
  The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
  The first three cases follow essentially the same argument provided by \citet{Wadler2022}, but we still give the arguments here for the sake of completeness.
  The last case \Rule{Op} is new and has been discussed above.
  For brevity we omit the subscript $(\Sigma, \Omega)$.
  \begin{itemize}
    \item \SynRule{Var}: If $\erase{\Gamma} |- \isTerm{t}^{\syn}$, then $(x : A) \in \Gamma$ and thus $\Gamma |- \isTerm{x} \syn A$.

@@

The formal counterpart of the above proof in \Agda functions as two top-level programs for type checking and synthesis.
These programs either compute the typing derivation or provide a proof of contradiction.
Each case analysis simply branches depending on the outcomes of bidirectional type synthesis and checking for each subterm, as well as the unification process.
If a contradiction proof is not of interest for implementation, these programs can be simplified by disregarding the cases that yield such contradiction proofs.
Alternatively, we could consider generalising typing derivations instead, like our generalised mode derivations (\Cref{fig:generalised-mode-derivations}).
This could accommodate ill-typed cases and reformulate contradiction proofs positively to deliver more informative error messages.
This would assist programmers in resolving issues with ill-typed terms, rather than simply returning a blatant 'no'.

@@

\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
  For any mode-correct bidirectional type system $(\Sigma, \Omega)$, 
  exactly one of the following holds:
  \begin{enumerate}
    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type $A$.
    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but\/ $\Gamma |/-_{\Sigma, \Omega} \isTerm{t} : A$.
    \item $\erase{\Gamma} |/-_{\Sigma, \Omega} \isTerm{t}^{\syn}$.
  \end{enumerate}
\end{corollary}
\begin{proof}
  Combine \Cref{lem:soundness-completeness,thm:mode-decoration} with \Cref{thm:bidirectional-type-synthesis-checking}.

\end{proof}

>>>>

\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
This section focuses on defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
We start with \Cref{sec:mode-correctness} by defining mode-correctness and showing the uniqueness of synthesised types.
This uniqueness means that any two synthesised types for the same raw term $t$ under the same context $\Gamma$ have to be equal.
It will be used especially in \Cref{subsec:bidirectional-synthesis-checking} for the proof of the decidability of bidirectional type synthesis and checking.
Then, we conclude this section with the trichotomy on raw terms in \Cref{subsec:trichotomy}.

\subsection{Mode Correctness}\label{sec:mode-correctness}
As \citet{Dunfield2021} outlined, mode-correctness for a bidirectional typing rule means that 
\begin{enumerate*}
\item each `input' type variable in a premise must be an `output' variable in `earlier' premises, or provided by the conclusion if the rule is checking;
\item each `output' type variable in the conclusion should be some `output' variable in a premise if the rule is synthesising.
\end{enumerate*}
Here `input' variables refer to variables in an extension context and in the type of a premise in checking mode.
It is important to note that the order of premises in a bidirectional typing rule also matters, since synthesised type variables are instantiated incrementally during type synthesis.

@@

Now let us define mode-correctness rigorously.
As we have outlined, the condition of mode-correctness for a synthesising rule is different from that of a checking rule, and the argument order also matters.
Defining the condition directly for a rule, and thus in our setting for an operation, can be somewhat intricate.
Instead, we choose to define the conditions for the argument list---more specifically, triples $\biargvec$ of an extension context $\Delta_i$, a type $A_i$, and a mode $\dir{d_i}$---pertaining to an operation, for an operation, and subsequently for a signature.
We also need some auxiliary definitions for the subset of variables of a type and of a context and the set of variables that have been synthesised:
\begin{definition}
  The finite subset\footnote{%
  There are various definitions for finite subsets of a set within Martin-L\"{o}f type theory.
  However, for our purposes, the choice among these definitions is not a matter of concern.}
  of \emph{(free) variables} of a type $A$ is denoted by $\fv(A)$.
  The subset $\fv(\Gamma)$ of variables in an extension context~$\Delta$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Delta, A) = \fv(\Delta) \cup \fv(A)$.
  For an argument list $\biargs$, define its set of \emph{synthesised type variables} inductively by 
  \begin{align}
    \label{eq:synvar1}\synvar(\cdot)                   & = \emptyset  \\
    \label{eq:synvar2}\synvar(\biargs, \chkbiarg[n+1]) & = \phantom{\fv(A_{n+1}) \cup {}} \synvar(\biargs) \\
    \label{eq:synvar3}\synvar(\biargs, \synbiarg[n+1]) & = \fv(A_{n+1}) \cup           \synvar(\biargs).
  \end{align}
\end{definition}
This subset contains type variables of a synthesising argument and they are exactly those type variables that will be synthesised during type synthesis.

\begin{definition}\label{def:mode-correctness-args}
  The \emph{mode-correctness} $\MCas\left(\biargvec\right)$ for an argument list $\biargs$ with respect to a subset $S$ of $\Xi$ is defined as
  \begin{align}
    \label{eq:MC1} \MCas( \cdot ) & = \top \\
    \label{eq:MC3} \MCas\left(\biargvec, \chkbiarg[n]\right)
                                  & = \fv(\Delta, A_n) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land \MCas\left(\biargvec\right) \\
    \label{eq:MC2} \MCas\left(\biargvec, \synbiarg[n]\right) 
                                  & = \phantom{, A_n} \fv(\Delta) \subseteq \left( S \cup \synvar\left(\biargvec\right)\right) \land  \MCas\left(\biargvec\right)
  \end{align}
  where \eqref{eq:MC1} means an empty list is always mode-correct.
\end{definition}
This definition encapsulates the idea that every `input' type variable, possibly derived from a context $\Delta$ or a checking argument $A_n$, must be an `output' variable from $\synvar(\biargvec)$ or, if the rule is checking, belong to the set $S$ of `input' variables in its conclusion.
This condition must be met for every tail of the argument list as well to ensure that `output' variables accessible at each argument position are from preceding arguments only, hence an inductive definition.
\begin{definition}\label{def:mode-correctness}
  An operation $\biop$ is \emph{mode-correct} if 
  \begin{enumerate}
    \item either $d$ is $\chk$, its argument list is mode-correct w.r.t.\ $\fv(A_0)$, and the union $\fv(A_0) \cup \synvar(\biargvec)$ contains every inhabitant of $\Xi$;
    \item or $d$ is $\syn$, its argument list is mode-correct w.r.t.\ $\emptyset$, and the finite subset $\synvar(\biargvec)$ contains every inhabitant of $\Xi$ and, particularly, $\fv(A_0)$.
  \end{enumerate}
  A bidirectional binding signature $\Omega$ is \emph{mode-correct} if its operations are all mode-correct.
\end{definition}
For a checking operation, an `input' variable of an argument could be derived from $A_0$, as these are known during type checking as an input.
Since every inhabitant of $\Xi$ can be located in either $A_0$ or synthesised variables, we can determine a concrete type for each inhabitant of $\Xi$ during type synthesis.
On the other hand, for a synthesising operation, we do not have any known variables at the onset of type synthesis, so the argument list should be mode-correct with respect to $\emptyset$.
Also, the set of synthesised variables alone should include every type variable in $\Xi$ and particularly in $A_n$.
It is also easy to check the bidirectional type system $(\Sigma_{\bto}, \Omega^{\Leftrightarrow}_{\Lambda}$) in \Cref{ex:signature-simply-typed-lambda} for simply typed $\lambda$-calculus is mode-correct according to our definition.

\begin{remark}
  Mode-correctness is fundamentally a condition for bidirectional typing \emph{rules} themselves, not for derivations.
  Thus, this property cannot be formalised without treating rules as some mathematical object, such as the notion of bidirectional binding signature presented in \Cref{sec:defs}.
  This contrasts with the properties in \Cref{sec:pre-synthesis}, which can still be specified for individual systems even in the absence of a generic definition of bidirectional type systems.
\end{remark}

Now, we set out to show the uniqueness of synthesised types for a mode-correct bidirectional type system.
For a specific system, its proof is typically a straightforward induction.
However, given the mode-correctness---which incorporates an inductive definition on the argument list---we establish our proof through induction, both on typing derivations and on the argument list:
\begin{lemma}[Uniqueness of Synthesised Types]\label{thm:unique-syn}
  In a mode-correct bidirectional type system $(\Sigma, \Omega)$, the synthesised types of any two derivations
  \[
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A
    \quad\text{and}\quad
    \Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B
  \]
  for the same term $t$ must be equal, i.e.\ $A = B$.
\end{lemma}
\begin{proof}%[Proof of \Cref{thm:unique-syn}]
  We prove the statement by induction on derivations $d_1$ and $d_2$ for $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn A$ and $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$.
  Our system is syntax-directed, so $d_1$ and $d_2$ must be derived from the same rule: 
  \begin{itemize}
    \item $\SynRule{Var}$ follows from that each variable as a raw term refers to the same variable in its context.
    \item $\SynRule{Anno}$ holds trivially, since the synthesised type $A$ is from the term $t \annotate A$ in question.
    \item $\Rule{Op}$: Recall that a derivation of\/ $\Gamma |- \tmOpts \syn A$ contains a substitution $\rho$ from the local context $\Xi$ to concrete types.
      To prove that any two typing derivations has the same synthesised type, it suffices to show that those substitutions $\rho_1$~and~$\rho_2$ of $d_1$~and~$d_2$, respectively, agree on variables in $\synvar(\biargs)$ so that $\simsub{A_0}{\rho_1} = \simsub{A_0}{\rho_2}$.
      We prove it by induction on the argument list:
      \begin{enumerate}
        \item For the empty list, the statement is vacuously true by \Cref{eq:synvar1}.
        \item If $\dir{d_{i+1}}$ is~$\chk$, then the statement holds for \Cref{eq:synvar2} by induction hypothesis.
        \item If $\dir{d_{i+1}}$ is~$\syn$, then $\simsub{\Delta_{i+1}}{\rho_1} = \simsub{\Delta_{i+1}}{\rho_2}$ by mode-correctness \Cref{eq:MC2} and induction hypothesis (of the list).
          Therefore, under the same context $\Gamma, \simsub{\Delta_{i+1}}{\rho_1} = \Gamma, \simsub{\Delta_{i+1}}{\rho_2}$ the term $t_{i+1}$ must have the same synthesised type $\simsub{A_{i+1}}{\rho_1} = \simsub{A_{i+1}}{\rho_1}$ by induction hypothesis (of the typing derivation), so $\rho_1$ and $\rho_2$ agree on $\fv(A_{i+1})$ in addition to $\synvar(\biargs)$, as required for \Cref{eq:synvar3} in the definition of $\synvar$.
      \end{enumerate}
    \vspace{-\topsep-\baselineskip}
  \end{itemize}
\end{proof}

@@

We have arrived at the main technical contribution of this paper.

@@

The interesting part of the theorem is the case for the $\Rule{Op}$ rule and it is rather complex.
Here we shall give its insight first instead of jumping into the details.
Recall that a typing derivation for $\tmOpts$ contains a substitution $\rho\colon \Xi \to \Type_{\Sigma}(\emptyset)$.
The goal of type synthesis is exactly to define such a substitution $\rho$, so we have to start with an `accumulating' substitution: a substitution $\rho_0$ that is partially defined on $\fv(A_0)$ if $d$ is $\chk$ or otherwise nowhere.
By mode-correctness, the accumulating substitution~$\rho_i$ will be defined on enough synthesised variables so that type synthesis or checking can be performed on $t_{i}$ with the context $\Gamma, \vars{x}_{i} : \simsub{\Delta_{i}}{\rho_{i}}$ based on its mode derivation $\erase{\Gamma}, \vars{x}_i |-_{\Sigma, \Omega} t_i^{\dir{d_i}}$.
If we visit a synthesising argument $\synbiarg[i+1]$, then we may extend the domain of $\rho_i$ to include the synthesised variables $\fv(A_{i + 1})$ if type synthesis is successful and also that the synthesised type can be \emph{unified with $A_{i+ 1}$} and thereby \emph{extend} $\rho_i$ to $\bar{\rho_i} = \rho_{i+1}$ with the unifier.
Then, if we go through every $t_i$ successfully, we will have a total substitution $\rho_n$ by mode-correctness and a derivation of $\Gamma, \vars{x}_i : \Delta_i |-_{\Sigma, \Omega} t_i :^{\dir{d_i}} \simsub{A}{\rho_n}$ for each subterm $t_i$.

@@

\begin{proof}[Proof of {\Cref{thm:bidirectional-type-synthesis-checking}}]
  We prove this statement by induction on the mode derivation\/ $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\dir{d}}$.
  The two cases \SynRule{Var} and \SynRule{Anno} are straightforward and have nothing to do with mode-correctness.
  The case \ChkRule{Sub} invokes the uniqueness of synthesised types to refute the case that $\Gamma |-_{\Sigma, \Omega} \isTerm{t} \syn B$ but $A \neq B$ for a given type $A$.
  The first three cases follow essentially the same reasoning provided by \citet{Wadler2022}, but we still present the reasoning here for the sake of completeness.
  The last case \Rule{Op} is new and has been discussed above.
  For brevity we omit the subscript $(\Sigma, \Omega)$.
  \begin{itemize}
    \item \SynRule{Var}: If $\erase{\Gamma} |- \isTerm{t}^{\syn}$, then $(x : A) \in \Gamma$ and thus $\Gamma |- \isTerm{x} \syn A$.

@@

The formal counterpart of the above proof in \Agda functions as two top-level programs for type checking and synthesis.
These programs either compute the typing derivation or provide a proof of contradiction.
Each case analysis simply branches depending on the outcomes of bidirectional type synthesis and checking for each subterm, as well as the unification process.
If a contradiction proof is not of interest for implementation, these programs can be simplified by disregarding the cases that yield such contradiction proofs.
Alternatively, we could consider generalising typing derivations instead, like our generalised mode derivations (\Cref{fig:generalised-mode-derivations}).
This could accommodate ill-typed cases and reformulate contradiction proofs positively to deliver more informative error messages.
This would assist programmers in resolving issues with ill-typed terms, rather than simply returning a blatant `no'.

@@

Combining the bidirectional synthesiser in \cref{subsec:bidirectional-synthesis-checking} with the mode decorator, soundness, and completeness in \cref{sec:pre-synthesis}, we can now construct a type synthesiser, which generalises \cref{thm:implementation}.

\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
  For any mode-correct bidirectional type system $(\Sigma, \Omega)$, 
  exactly one of the following holds:
  \begin{enumerate}
    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ and\/ $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ for some type~$A$.
    \item $\erase{\Gamma} |-_{\Sigma, \Omega} \isTerm{t}^{\syn}$ but\/ $\Gamma |/-_{\Sigma, \Omega} \isTerm{t} : A$ for any type~$A$.
    \item $\erase{\Gamma} |/-_{\Sigma, \Omega} \isTerm{t}^{\syn}$.
  \end{enumerate}
\end{corollary}
\begin{proof}
  Combine \cref{thm:bidirectional-type-synthesis-checking} with \cref{lem:soundness-completeness,thm:mode-decoration}.
\end{proof}
